Basics of cubic spline models

### 2021-11-23 text/21/cubic-splines.Rmd
Added b-splines to cubic-splines.Rmd

```{r setup}
suppressMessages(suppressWarnings(library(broom)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(magrittr)))
```

@@

```{r}
library(ggplot2)
x <- 1:20
y <- rep(40, 20)
y[6:20] <- y[6:20]-3*(1:15)
y[11:20] <- y[11:20]+7*(1:10)
y[16:20] <- y[16:20]-(1:5)^2
y <- y+round(rnorm(10), 1)
simulated_example <- data.frame(x, y)
simulated_example
ggplot(simulated_example, aes(x, y)) +
  geom_point()
```

A single cubic function for all the data doesn't quite capture what is happening here.

```{r}
extra_values <- data.frame(x=seq(0, 21, length=2200))
simulated_example %<>% full_join(extra_values, by="x")
```

```{r}
single_cubic_model <- lm(y~poly(x, 3), data=simulated_example)
single_cubic_prediction <- augment(single_cubic_model)
ggplot(single_cubic_prediction, aes(x, y)) + 
  geom_point() +
  geom_line(aes(y=.fitted))
```

@@

```{r}
sb1 <- (simulated_example$x<=5)
sb2 <- (simulated_example$x>5)&(simulated_example$x<=10)
sb3 <- (simulated_example$x>10)&(simulated_example$x<=15)
sb4 <- (simulated_example$x>15)

simulated_example %>%
  filter(sb1) %>%
  lm(y~poly(x, 3), data=.) %>%
  augment(newdata=simulated_example[sb1, ]) -> first_cubic_prediction
lm(y~poly(x, 3), subset=sb2, data=simulated_example) %>% 
  augment(newdata=simulated_example[sb2, ]) %>%
  select(x, .fitted) -> second_cubic_prediction
lm(y~poly(x, 3), subset=sb3, data=simulated_example) %>%
  augment(newdata=simulated_example[sb3, ]) %>% 
  select(x, .fitted) -> third_cubic_prediction
lm(y~poly(x, 3), subset=sb4, data=simulated_example) %>%
  augment(newdata=simulated_example[sb4, ]) %>%
  select(x, .fitted) -> fourth_cubic_prediction

first_cubic_prediction %>%
  bind_rows(second_cubic_prediction) %>%
  bind_rows(third_cubic_prediction) %>%
  bind_rows(fourth_cubic_prediction) %>%
  full_join(simulated_example, by="x") -> combined_prediction
ggplot(combined_prediction, aes(x, y.y)) +
  geom_point() +
  geom_line(aes(y=.fitted))
```

This function is not continuous or smooth. To make a single smooth continuous function you need to define

```{r}
simulated_example %>%
  mutate(x1=x) %>%
  mutate(x2=x^2) %>%
  mutate(x3=x^3) %>%
  mutate(x4=(x>5)*(x-5)^3) %>%
  mutate(x5=(x>10)*(x-10)^3) %>%
  mutate(x6=(x>15)*(x-15)^3) -> spline_basis
```

```{r}
lm(y~x1+x2+x3+x4+x5+x6, data=spline_basis) %>%
  augment(newdata=spline_basis) %>%
  ggplot(aes(x, y)) +
    geom_point() +
  geom_line(aes(y=.fitted))
```

>>>>

```{r setup}
suppressMessages(suppressWarnings(library(broom)))
suppressMessages(suppressWarnings(library(dplyr)))
suppressMessages(suppressWarnings(library(knitr)))
suppressMessages(suppressWarnings(library(magrittr)))
suppressMessages(suppressWarnings(library(splines)))
suppressMessages(suppressWarnings(library(tidyr)))
```

@@

```{r}
library(ggplot2)
x <- 0:20
y <- rep(40, 21)
y[6:21] <- y[6:21]-3*(1:15)
y[11:21] <- y[11:21]+7*(1:10)
y[16:21] <- y[16:21]-(1:5)^2
y <- y+round(rnorm(10), 1)
simulated_example <- data.frame(x, y)
simulated_example
ggplot(simulated_example, aes(x, y)) +
  geom_point()
```

To make the graphs look nice, you need to include some intermediate values.

```{r}
x <- c(0:20, setdiff(seq(0, 20, by=1/64), 0:20))
xm <- data.frame(
  c1 =rep(1, length(x)),
  c2 =x,
  c3 =x^2,
  c4 =x^3,
  c5 =(x> 5)*rep(1, length(x)),
  c6 =(x> 5)*(x- 5),
  c7 =(x> 5)*(x- 5)^2,
  c8 =(x> 5)*(x- 5)^3,
  c9 =(x>10)*rep(1, length(x)),
  c10=(x>10)*(x-10),
  c11=(x>10)*(x-10)^2,
  c12=(x>10)*(x-10)^3,
  c13 =(x>15)*rep(1, length(x)),
  c14=(x>15)*(x-15),
  c15=(x>15)*(x-15)^2,
  c16=(x>15)*(x-15)^3,
  y=NA
)
xm$y[1:21] <- y
```

```{r}
kable(xm[1:21, 1:4])
single_cubic_model <- lm(y~c2+c3+c4, data=xm)
single_cubic_model %>% 
  augment(newdata=xm) %>%
  ggplot(aes(x, y)) + 
    geom_point() +
    geom_line(aes(y=.fitted)) +
    ggtitle("Single cubic fit to the data")
```

@@

The trick to fitting four separate cubic polynomials is to "restart" the cubic terms after x=5, x=10, and x=15.

```{r}
kable(xm[1:20, ])
xm$y <- NA
xm$y[1:20] <- y
four_independent_cubics <- 
  lm(y~c2+c3+c4+c5+c6+c7+c8+c9+c10+c11+c12+c13+c14+c15+c16, data=xm)
augment(four_independent_cubics, newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y)) + 
    ggtitle("Discontinous cubic splines")
```

This function is not continuous or smooth. To make the function continuous, drop the extra intercept terms.

```{r}
kable(xm[1:20, c(1:4, 6:8, 10:12, 14:16)])
four_continuous_cubics <- 
  lm(y~c2+c3+c4+c6+c7+c8+c10+c11+c12+c14+c15+c16, data=xm)
augment(four_continuous_cubics, newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y)) +
    ggtitle("Continous but not smooth cubic splines")

```

Smoothness is measured in terms of the continuity of derivatives. Here is a function that has a continuous first derivative.

```{r}
kable(xm[1:20, c(1:4, 7:8, 11:12, 15:16)])
four_smooth_cubics <- 
  lm(y~c2+c3+c4+c7+c8+c11+c12+c15+c16, data=xm)
augment(four_smooth_cubics, newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y)) + 
    ggtitle("Continous and Smooth (1st derivative) splines")

```

Here is a function that has continuous first and second derivatives.

```{r}
kable(xm[1:20, c(1:4, 8, 12, 16)])
spline_model <- 
  lm(y~c2+c3+c4+c8+c12+c16, data=xm)
augment(spline_model, newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y)) +
    ggtitle("Continous and smooth (1st and 2nd derivatives) cubic splines")
```

While this approach is simple, there is an issue with multicollinearity.

```{r}
round(cor(xm[c(2:4, 8, 12, 16)]), 2)
```

This approach is simple and easy to follow, but there is one catch. The correlations are quite high and this can lead to computational problems, including rounding errors. So most spline models implemented on a computer use a different approach.

B-splines provide a solution with less issues of multi-collinearity.

```{r}
linear_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=1, intercept=TRUE)
kable(linear_spline[1:21, ])
plot(x[1:21], linear_spline[1:21 , 1])
plot(x[1:21], linear_spline[1:21 , 2])
plot(x[1:21], linear_spline[1:21 , 3])
plot(x[1:21], linear_spline[1:21 , 4])
plot(x[1:21], linear_spline[1:21 , 5])

lm(xm$y~linear_spline) %>%
  augment(newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y)) + 
    ggtitle("Linear splines")
```

```{r}
quadratic_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=2, intercept=TRUE)
kable(quadratic_spline[1:21, ])
plot(x[1:21], quadratic_spline[1:21 , 1])
plot(x[1:21], quadratic_spline[1:21 , 2])
plot(x[1:21], quadratic_spline[1:21 , 3])
plot(x[1:21], quadratic_spline[1:21 , 4])
plot(x[1:21], quadratic_spline[1:21 , 5])

lm(xm$y~quadratic_spline) %>%
  augment(newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y)) + 
    ggtitle("quadratic splines")
```

```{r}
cubic_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=3, intercept=TRUE)
kable(cubic_spline[1:21, ])
plot(x[1:21], cubic_spline[1:21 , 1])
plot(x[1:21], cubic_spline[1:21 , 2])
plot(x[1:21], cubic_spline[1:21 , 3])
plot(x[1:21], cubic_spline[1:21 , 4])
plot(x[1:21], cubic_spline[1:21 , 5])

lm(xm$y~cubic_spline) %>%
  augment(newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y)) + 
    ggtitle("cubic splines")
```
